---
title: "Macrophage"
author: "Summer"
date: "2025-06-12"
output: html_document
---
--------------------
Key
--------------------
id: id number of each experiment <chr>
gene_type: different types of RNAs and pseudo genes <chr>
gene_name: different gene names that re being transcribed <chr>
padj: adjusted p value (of comparison) <double>
wt0wt6: WT @ 0h vs WT @ 6h <double>
wt0wt18: WT @ 0h vs WT @ 18h <double>
ko0ko6: KO @ 0h vs KO @ 6h <double>
ko0ko18: KO @ 0h vs KO @ 18h <double>
ko0wt0: KO @ 0h vs WT @ 0h <double>
ko0wt6: KO @ 0h vs WT @ 6h <double>
ko0wt18: KO @ 0h vs WT @ 18h <double>

--------------------
Groups In Study 
--------------------
WT BMD macrophage (Tc negative / o hrs)
WT BMD macrophage (Tc positive, incubation time 6hrs)
WT BMD macrophage (Tc positive, incubation time 18hrs)
PARP KO (PARP gene knock out) BMD macrophage (Tc negative / o hrs)
PARP KO BMD macrophage (Tc positive, incubation time 6hrs)
PARP KO BMD macrophage (Tc positive, incubation time 18hrs)

--------------------
Definitions
--------------------
WT: Wild type - Parp1 is present
KO: Knock out - Parp1 has been silenced
BMD: 'bone marrow derived'
Macrophage: type of white blood cells that surrounds and kills microorganisms, removes dead cells, and stimulates the action of other immune system cells
Parp1: nuclear enzyme that plays a role in DNA repair, chromatin structure, and gene expression
Tc: parasite (parasitic protozoan) responsible for chagas disease

--------------------
Install important packages and libraries
--------------------
```{r package}
library(dplyr)
library(readxl)
library(corrplot)
library(ggplot2)
library(tidyverse)
library(emmeans)
library(ggpubr)
library(gridExtra)
library(tidyr)
library(broom)
library(ggrepel)
library(writexl)
library(rlang)
```

--------------------
Load data from excel
--------------------
```{r load}
file_path <- ("/Users/summerjohnson/Documents/GitHub/UTMB-SIBDS/macrophagextcNGS.xlsx")
data <- read_xlsx("/Users/summerjohnson/Documents/GitHub/UTMB-SIBDS/macrophagextcNGS.xlsx")
```

--------------------
Column Handling
--------------------
```{r redo}
#Rename columns
colnames(data) [colnames(data) == "ENSEMBL_id"] <- "id"
colnames(data) [colnames(data) == "type"] <- "gene_type"
colnames(data) [colnames(data) == "name"] <- "gene_name"
colnames(data) [colnames(data) == "WT vs WT @ 6h"] <- "wt0wt6"
colnames(data) [colnames(data) == "padj...5"] <- "padj_wt0wt6"
colnames(data) [colnames(data) == "WT vs WT @ 18h"] <- "wt0wt18"
colnames(data) [colnames(data) == "padj...8"] <- "padj_wt0wt18"
colnames(data) [colnames(data) == "KO vs KO @ 6h"] <- "ko0ko6"
colnames(data) [colnames(data) == "padj...11"] <- "padj_ko0ko6"
colnames(data) [colnames(data) == "KO vs KO @ 18h"] <- "ko0ko18"
colnames(data) [colnames(data) == "padj...14"] <- "padj_ko0ko18"
colnames(data) [colnames(data) == "KO vs WT @ 0h"] <- "ko0wt0"
colnames(data) [colnames(data) == "padj...17"] <- "padj_ko0wt0"
colnames(data) [colnames(data) == "KO vs WT @ 6h"] <- "ko0wt6"
colnames(data) [colnames(data) == "padj...20"] <- "padj_ko0wt6"
colnames(data) [colnames(data) == "KO vs WT @ 18h"] <- "ko0wt18"
colnames(data) [colnames(data) == "padj...23"] <- "padj_ko0wt18"

#Create subset with only necessary columns
tc_data_1 <- data[, c("id", "gene_type", "gene_name",
                    "wt0wt6", "padj_wt0wt6",
                    "wt0wt18", "padj_wt0wt18",
                    "ko0ko6", "padj_ko0ko6",
                    "ko0ko18", "padj_ko0ko18",
                    "ko0wt0", "padj_ko0wt0",
                    "ko0wt6", "padj_ko0wt6",
                    "ko0wt18", "padj_ko0wt18")]
```

--------------------
NA Handling
- Dataset with rows without information (NA) removed
- NA represents where information was absent
--------------------
``` {r na}
#Remove NA variables from comparison columns
tc_data <- tc_data_1 %>%
  filter(!(wt0wt6 == "NA" & wt0wt18 == "NA" & ko0ko6 == "NA" & ko0ko18 == "NA"
           & ko0wt0 == "NA" & ko0wt6 == "NA" & ko0wt18 == "NA"))

#Convert character columns to numeric
cols_to_convert <- setdiff(colnames(tc_data), c("id", "gene_type", "gene_name" ))

tc_data[cols_to_convert] <- lapply(tc_data[cols_to_convert], as.numeric)

```

--------------------
EDA - User Input
- Histograms of different comparison scores to check for normalization
--------------------
```{r eda}
#User input for histogram
print("Comparison Type Options: 'wt0wt6', 'wt0wt18', 'ko0ko6', 'ko0ko18', 'ko0wt0', 'ko0wt6', 'ko0wt18'")
comparison_type_eda <- readline(prompt = "Enter which comparison you would like to see the significant increases of: ")

# Generate the histogram based on user input
if (comparison_type_eda %in% names(tc_data)) {
  ggplot(tc_data, aes(x = tc_data[[comparison_type_eda]])) +
    geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
    labs(
      title = paste("Histogram of", comparison_type_eda),
      x = comparison_type_eda,
      y = "Count"
    )
} else {
  message("Invalid comparison type entered.")
}

```

--------------------
Volcano Plots - User Input
- Assess the relationship between magnitude of change and statistical significance of genes
- Large changes and higher significance towards the top and edges of the plot
--------------------
``` {r volcano}
# Define valid comparisons and their corresponding p-value columns
valid_comparisons <- c("wt0wt6", "wt0wt18", "ko0ko6", "ko0ko18", "ko0wt0", "ko0wt6", "ko0wt18")
padj_lookup <- list(
  wt0wt6 = "padj_wt0wt6",
  wt0wt18 = "padj_wt0wt18",
  ko0ko6 = "padj_ko0ko6",
  ko0ko18 = "padj_ko0ko18",
  ko0wt0 = "padj_ko0wt0",
  ko0wt6 = "padj_ko0wt6",
  ko0wt18 = "padj_ko0wt18"
)

# Ask user for comparison
print(paste("Comparison Type Options:", paste(valid_comparisons, collapse = ", ")))
comparison_type_volc <- readline(prompt = "Enter which comparison you would like to see the volcano plot of: ")

# Check validity and plot
if (comparison_type_eda %in% valid_comparisons) {
  x_col <- comparison_type_volc
  y_col <- padj_lookup[[comparison_type_volc]]
  
  ggplot(tc_data, aes(x = tc_data[[x_col]], y = -log10(tc_data[[y_col]]))) +
    geom_point(color = "steelblue") +
    labs(
      title = paste("Volcano Plot of", comparison_type_volc),
      x = x_col,
      y = "-log10(padj)"
    )
  
} else {
  message("Invalid comparison type entered.")
}

```

--------------------
Sig Inc by Gene Type Using User input
--------------------
``` {r user_input_increase}
# Prompt user for inputs
print("Gene Type Options: ")
unique(tc_data$gene_type)
print("Comparison Type Options: 'wt0wt5', 'wt0wt18', 'ko0ko6', 'ko0ko18', 'ko0wt0', 'ko0wt6', 'ko0wt18'")
rna_type <- readline(prompt = "Enter which gene type you would like to see the significant increases of: ")
comparison_type <- readline(prompt = "Enter which comparison you would like to see the significant increases of: ")

# Dynamically refer to columns
comparison_col <- sym(comparison_type)
padj_col <- sym(paste0("padj_", comparison_type))

# Filter and create subset
sig_subset_inc <- tc_data %>%
  filter(
    gene_type == rna_type,
    !!padj_col < 0.05,
    !!comparison_col > 0
  ) %>%
  select(gene_type, gene_name, !!comparison_col, !!padj_col) %>%
  arrange(!!comparison_col)

# Rename selected columns for consistent plotting
colnames(sig_subset_inc)[which(names(sig_subset_inc) == comparison_type)] <- "comparison_value"
colnames(sig_subset_inc)[which(names(sig_subset_inc) == paste0("padj_", comparison_type))] <- "padj_value"

# Create plot
sig_inc_plot <- ggplot(sig_subset_inc, aes(x = reorder(gene_name, comparison_value), y = comparison_value, fill = gene_name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 7),
    plot.margin = margin(10, 10, 10, 30)
  ) +
  labs(
    x = "Gene",
    y = paste0("Change in ", comparison_type),
    title = paste("Significant Increase in", rna_type, "Expression for", comparison_type)
  )

# Save plot
plot_filename <- paste0(rna_type, "_", comparison_type, "_sig_inc_bp.png")
ggsave(plot_filename, plot = sig_inc_plot, bg = "white", width = 8, height = 12, dpi = 300)

# Save Excel file
excel_filename <- paste0("sig_", rna_type, "_", comparison_type, "_inc.xlsx")
write_xlsx(sig_subset_inc, excel_filename)
#-----------------
```

--------------------
Sig Dec by Gene Type Using User input
--------------------
``` {r user_input_decrease}
# Prompt user for inputs
print("Gene Type Options: ")
unique(tc_data$gene_type)
print("Comparison Type Options: 'wt0wt5', 'wt0wt18', 'ko0ko6', 'ko0ko18', 'ko0wt0', 'ko0wt6', 'ko0wt18'")
rna_type <- readline(prompt = "Enter which gene type you would like to see the significant decreases of: ")
comparison_type <- readline(prompt = "Enter which comparison you would like to see the significant decreases of: ")

# Dynamically refer to columns
comparison_col <- sym(comparison_type)
padj_col <- sym(paste0("padj_", comparison_type))

# Filter and create subset
sig_subset_dec <- tc_data %>%
  filter(
    gene_type == rna_type,
    !!padj_col < 0.05,
    !!comparison_col < 0
  ) %>%
  select(gene_type, gene_name, !!comparison_col, !!padj_col) %>%
  arrange(!!comparison_col)

# Rename selected columns for consistent plotting
colnames(sig_subset_dec)[which(names(sig_subset_dec) == comparison_type)] <- "comparison_value"
colnames(sig_subset_dec)[which(names(sig_subset_dec) == paste0("padj_", comparison_type))] <- "padj_value"

# Create plot
sig_dec_plot <- ggplot(sig_subset_dec, aes(x = reorder(gene_name, comparison_value), y = comparison_value, fill = gene_name)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 7),
    plot.margin = margin(10, 10, 10, 30)
  ) +
  labs(
    x = "Gene",
    y = paste0("Change in ", comparison_type),
    title = paste("Significant Decrease(s) in", rna_type, "Expression for", comparison_type)
  )

# Save plot
plot_filename <- paste0(rna_type, "_", comparison_type, "_sig_dec_bp.png")
ggsave(plot_filename, plot = sig_dec_plot, bg = "white", width = 8, height = 12, dpi = 300)

# Save Excel file
excel_filename <- paste0("sig_", rna_type, "_", comparison_type, "_dec.xlsx")
write_xlsx(sig_subset_dec, excel_filename)
#-----------------
```

--------------------
Significant Magnitudes of Change
--------------------
``` {r significant}
#Threshold
padj_threshold <- 0.05

#Significant for all comparisons
sig_all <- tc_data %>%
  filter(
    padj_wt0wt6 < padj_threshold &
    padj_wt0wt18 < padj_threshold &
    padj_ko0ko6 < padj_threshold &
    padj_ko0ko18 < padj_threshold &
    padj_ko0wt0 < padj_threshold &
    padj_ko0wt6 < padj_threshold &
    padj_ko0wt18 < padj_threshold
  )

#Significant for all wt vs wt comparisons
sig_wt <- tc_data %>%
  filter(
    padj_wt0wt6 < padj_threshold &
    padj_wt0wt18 < padj_threshold 
  )

#Significant for all ko vs ko comparisons
sig_ko <- tc_data %>%
  filter(
    padj_ko0ko6 < padj_threshold &
    padj_ko0ko18 < padj_threshold 
  )

#Significant for all ko vs wt comparisons
sig_wtko <- tc_data %>%
  filter(
    padj_ko0wt0 < padj_threshold &
    padj_ko0wt6 < padj_threshold &
    padj_ko0wt18 < padj_threshold
  )

#Individual significance
sig_wt0wt6 <- tc_data %>%
  filter(padj_wt0wt6 < padj_threshold)

sig_wt0wt18 <- tc_data %>%
  filter(padj_wt0wt18 < padj_threshold)

sig_ko0ko6 <- tc_data %>%
  filter(padj_ko0ko6 < padj_threshold)

sig_ko0ko18 <- tc_data %>%
  filter(padj_ko0ko18 < padj_threshold)

sig_ko0wt0 <- tc_data %>%
  filter(padj_ko0wt0 < padj_threshold)

sig_ko0wt6 <- tc_data %>%
  filter(padj_ko0wt6 < padj_threshold)

sig_ko0wt18 <- tc_data %>%
  filter(padj_ko0wt18 < padj_threshold)

```

--------------------
Repeating Genes
--------------------
``` {r compare}
#Keep only gene_names that have more than one gene_type
multi_type_genes <- tc_data %>%
  arrange(gene_name) %>%
  group_by(gene_name) %>%
  filter(n_distinct(gene_type) > 1) %>%
  ungroup() %>%
  arrange(gene_name)

print(multi_type_genes)
#Only 7 genes that repeat in different types (that have data collected)
#Only 1 gene (Zkscan7) that has adjusted p values for both types during comparisons

#--------------------
#Select relevant columns for wt0wt6
wt0wt6_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_wt0wt6) %>%
  filter(!is.na(padj_wt0wt6)) %>%
  arrange(gene_name, gene_type)

#Alter for wt0wt6 plot
multi_type_genes_1 <- multi_type_genes %>%
    mutate(significant = case_when(
      is.na(padj_wt0wt6) ~ "-",
      padj_wt0wt6 < 0.05 ~ "*",
      TRUE ~ NA_character_
    ),
         label_y = ifelse(wt0wt6 >= 0,
                          wt0wt6 + 0.2,
                          wt0wt6 - 0.2))
#wt0wt6 plot
wt0wt6_compplot <- ggplot(multi_type_genes_1, aes(x = gene_type, y = wt0wt6, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_1$wt0wt6 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "WT @ 0h vs WT @ 6h", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
wt0wt6_sig_by_type
wt0wt6_compplot
#--------------------

#--------------------
#Select relevant columns for wt0wt18
wt0wt18_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_wt0wt18) %>%
  filter(!is.na(padj_wt0wt18)) %>%
  arrange(gene_name, gene_type)

#Alter for wt0wt18 plot
multi_type_genes_2 <- multi_type_genes %>%
    mutate(significant = case_when(
    is.na(padj_wt0wt18) ~ "-",
    padj_wt0wt18 < 0.05 ~ "*",
    TRUE ~ NA_character_
    ),
         label_y = ifelse(wt0wt18 >= 0,
                          wt0wt18 + 0.2,
                          wt0wt18 - 0.2))
#wt0wt18 plot 
wt0wt18_compplot <- ggplot(multi_type_genes_2, aes(x = gene_type, y = wt0wt18, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_2$wt0wt18 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "WT @ 0h vs WT @ 18", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
wt0wt18_sig_by_type
wt0wt18_compplot
#--------------------

#--------------------
#Select relevant columns for ko0ko6
ko0ko6_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_ko0ko6) %>%
  filter(!is.na(padj_ko0ko6)) %>%
  arrange(gene_name, gene_type)

#Alter for ko0ko6 plot
multi_type_genes_3 <- multi_type_genes %>%
    mutate(significant = case_when(
    is.na(padj_ko0ko6) ~ "-",
    padj_ko0ko6 < 0.05 ~ "*",
    TRUE ~ NA_character_
    ),
         label_y = ifelse(ko0ko6 >= 0,
                          ko0ko6 + 0.2,
                          ko0ko6 - 0.2))
#ko0ko6 plot
ko0ko6_compplot <- ggplot(multi_type_genes_3, aes(x = gene_type, y = ko0ko6, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_3$ko0ko6 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "KO @ 0h vs KO @ 6h", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
ko0ko6_sig_by_type
ko0ko6_compplot
#--------------------

#--------------------
#Select relevant columns for ko0ko18
ko0ko18_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_ko0ko18) %>%
  filter(!is.na(padj_ko0ko18)) %>%
  arrange(gene_name, gene_type)

#Alter for ko0ko18 plot
multi_type_genes_4 <- multi_type_genes %>%
    mutate(significant = case_when(
    is.na(padj_ko0ko18) ~ "-",
    padj_ko0ko18 < 0.05 ~ "*",
    TRUE ~ NA_character_
    ),
         label_y = ifelse(ko0ko18 >= 0,
                          ko0ko18 + 0.2,
                          ko0ko18 - 0.2))
#ko0ko6 plot
ko0ko18_compplot <- ggplot(multi_type_genes_4, aes(x = gene_type, y = ko0ko18, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_4$ko0ko18 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "KO @ 0h vs KO @ 18h", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
ko0ko18_sig_by_type
ko0ko18_compplot
#--------------------

#--------------------
#Select relevant columns for ko0wt0
ko0wt0_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_ko0wt0) %>%
  filter(!is.na(padj_ko0wt0)) %>%
  arrange(gene_name, gene_type)

#Alter for ko0wt0 plot
multi_type_genes_5 <- multi_type_genes %>%
    mutate(significant = case_when(
    is.na(padj_ko0wt0) ~ "-",
    padj_ko0wt0 < 0.05 ~ "*",
    TRUE ~ NA_character_
    ),
         label_y = ifelse(ko0wt0 >= 0,
                          ko0wt0 + 0.2,
                          ko0wt0 - 0.2))
#ko0wt0 plot
ko0wt0_compplot <- ggplot(multi_type_genes_5, aes(x = gene_type, y = ko0wt0, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_5$ko0wt0 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "KO @ 0h vs WT @ 0h", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
ko0wt0_sig_by_type
ko0wt0_compplot
#--------------------

#--------------------
#Select relevant columns for ko0wt6
ko0wt6_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_ko0wt6) %>%
  filter(!is.na(padj_ko0wt6)) %>%
  arrange(gene_name, gene_type)

#Alter for ko0wt6 plot
multi_type_genes_6 <- multi_type_genes %>%
  mutate(significant = case_when(
    is.na(padj_ko0wt6) ~ "-",
    padj_ko0wt6 < 0.05 ~ "*",
    TRUE ~ NA_character_
    ),
         label_y = ifelse(ko0wt6 >= 0,
                          ko0wt6 + 0.2,
                          ko0wt6 - 0.2))
#ko0wt6 plot
ko0wt6_compplot <- ggplot(multi_type_genes_6, aes(x = gene_type, y = ko0wt6, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_6$ko0wt6 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "KO @ 0h vs WT @ 6h", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
ko0wt6_sig_by_type
ko0wt6_compplot
#--------------------

#--------------------
#Select relevant columns for ko0wt18
ko0wt18_sig_by_type <- multi_type_genes %>%
  select(gene_name, gene_type, padj_ko0wt18) %>%
  filter(!is.na(padj_ko0wt18)) %>%
  arrange(gene_name, gene_type)

#Alter for ko0wt18 plot
multi_type_genes_7 <- multi_type_genes %>%
  mutate(significant = case_when(
    is.na(padj_ko0wt18) ~ "-",
    padj_ko0wt18 < 0.05 ~ "*",
    TRUE ~ NA_character_
    ),
         label_y = ifelse(ko0wt18 >= 0,
                          ko0wt18 + 0.2,
                          ko0wt18 - 0.2))
#ko0wt18 plot
ko0wt18_compplot <- ggplot(multi_type_genes_7, aes(x = gene_type, y = ko0wt18, color = gene_type)) +
  geom_point(size = 4, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = significant, y = label_y), vjust = ifelse(multi_type_genes_7$ko0wt18 >= 0, 0, 1), size = 5) +
  facet_wrap(~gene_name, scale = "fixed") +
  theme_minimal() +
  labs(x = "Gene Type", y = "KO @ 0h vs WT @ 18h", title = "Gene Comparisons w/ Significance", subtitle = "* = p < 0.05, - = missing p-value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

#View
ko0wt18_sig_by_type
ko0wt18_compplot
#--------------------

```


