---
title: "Main"
author: "Summer"
date: "2025-06-12"
output: html_document
---
--------------------
Key
--------------------
id: id number of each experiment <chr>
gene_type: different types of RNAs and pseudo genes <chr>
gene_name: different gene names that re being transcribed <chr>
padj: adjusted p value (of comparison) <double>
wt0wt6: WT @ 0h vs WT @ 6h <double>
wt0wt18: WT @ 0h vs WT @ 18h <double>
ko0ko6: KO @ 0h vs KO @ 6h <double>
ko0ko18: KO @ 0h vs KO @ 18h <double>
ko0wt0: KO @ 0h vs WT @ 0h <double>
ko0wt6: KO @ 0h vs WT @ 6h <double>
ko0wt18: KO @ 0h vs WT @ 18h <double>

--------------------
Groups In Study 
--------------------
WT BMD macrophage (Tc negative / o hrs)
WT BMD macrophage (Tc positive, incubation time 6hrs)
WT BMD macrophage (Tc positive, incubation time 18hrs)
PARP KO (PARP gene knock out) BMD macrophage (Tc negative / o hrs)
PARP KO BMD macrophage (Tc positive, incubation time 6hrs)
PARP KO BMD macrophage (Tc positive, incubation time 18hrs)

--------------------
Definitions
--------------------
WT: Wild type - Parp1 is present
KO: Knock out - Parp1 has been silenced
BMD: 'bone marrow derived'
Macrophage: type of white blood cells that surrounds and kills microorganisms, removes dead cells, and stimulates the action of other immune system cells
Parp1: nuclear enzyme that plays a role in DNA repair, chromatin structure, and gene expression
Tc: parasite (parasitic protozoan) responsible for chagas disease

--------------------
Install important packages and libraries
--------------------
```{r package}
library(dplyr)
library(readxl)
library(corrplot)
library(ggplot2)
library(tidyverse)
library(emmeans)
library(ggpubr)
library(gridExtra)
```

--------------------
Load data from excel
--------------------
```{r load}
file_path <- ("/Users/summerjohnson/Documents/GitHub/UTMB-SIBDS/macrophagextcNGS.xlsx")
data <- read_xlsx("/Users/summerjohnson/Documents/GitHub/UTMB-SIBDS/macrophagextcNGS.xlsx")
```

--------------------
Column Handling
--------------------
```{r redo}
#Rename columns
colnames(data) [colnames(data) == "ENSEMBL_id"] <- "id"
colnames(data) [colnames(data) == "type"] <- "gene_type"
colnames(data) [colnames(data) == "name"] <- "gene_name"
colnames(data) [colnames(data) == "WT vs WT @ 6h"] <- "wt0wt6"
colnames(data) [colnames(data) == "padj...5"] <- "padj_wt0wt6"
colnames(data) [colnames(data) == "WT vs WT @ 18h"] <- "wt0wt18"
colnames(data) [colnames(data) == "padj...8"] <- "padj_wt0wt18"
colnames(data) [colnames(data) == "KO vs KO @ 6h"] <- "ko0ko6"
colnames(data) [colnames(data) == "padj...11"] <- "padj_ko0ko6"
colnames(data) [colnames(data) == "KO vs KO @ 18h"] <- "ko0ko18"
colnames(data) [colnames(data) == "padj...14"] <- "padj_ko0ko18"
colnames(data) [colnames(data) == "KO vs WT @ 0h"] <- "ko0wt0"
colnames(data) [colnames(data) == "padj...17"] <- "padj_ko0wt0"
colnames(data) [colnames(data) == "KO vs WT @ 6h"] <- "ko0wt6"
colnames(data) [colnames(data) == "padj...20"] <- "padj_ko0wt6"
colnames(data) [colnames(data) == "KO vs WT @ 18h"] <- "ko0wt18"
colnames(data) [colnames(data) == "padj...23"] <- "padj_ko0wt18"

#Create subset with only necessary columns
tc_data_1 <- data[, c("id", "gene_type", "gene_name",
                    "wt0wt6", "padj_wt0wt6",
                    "wt0wt18", "padj_wt0wt18",
                    "ko0ko6", "padj_ko0ko6",
                    "ko0ko18", "padj_ko0ko18",
                    "ko0wt0", "padj_ko0wt0",
                    "ko0wt6", "padj_ko0wt6",
                    "ko0wt18", "padj_ko0wt18")]
```

--------------------
NA Handling
- Dataset with rows without information (NA) removed
- NA represents where information was absent
--------------------
``` {r na}
#Remove NA variables from comparison columns
tc_data <- tc_data_1 %>%
  filter(!(wt0wt6 == "NA" & wt0wt18 == "NA" & ko0ko6 == "NA" & ko0ko18 == "NA"
           & ko0wt0 == "NA" & ko0wt6 == "NA" & ko0wt18 == "NA"))

#Convert character columns to numeric
cols_to_convert <- setdiff(colnames(tc_data), c("id", "gene_type", "gene_name" ))

tc_data[cols_to_convert] <- lapply(tc_data[cols_to_convert], as.numeric)

```

--------------------
EDA
- Histograms of different comparison scores to check for normalization
--------------------
```{r eda}
#Histogram of WT @ 0h vs WT @ 6h
hist_1 <- ggplot(tc_data, aes(x = wt0wt6)) +
  geom_histogram(binwidth = 1, fill = "aquamarine", color = "black") +
  labs(title = "Histogram of WT @ 0h vs WT @ 6h", x = "WT @ 0h vs WT @ 6h", y = "Count")

#Histogram of WT @ 0h vs WT @ 18h
hist_2 <- ggplot(tc_data, aes(x = wt0wt18)) +
  geom_histogram(binwidth = 1, fill = "coral", color = "black") +
  labs(title = "Histogram of WT @ 0h vs WT @ 18h", x ="WT @ 0h vs WT @ 18h", y = "Count")

#Histogram of KO @ 0h vs KO @ 6h
hist_3 <- ggplot(tc_data, aes(x = ko0ko6)) +
  geom_histogram(binwidth = 1, fill = "brown2", color = "black") +
  labs(title = "Histogram of KO @ 0h vs KO @ 6h", x = "KO @ 0h vs KO @ 6h", y = "Count")

#Histogram of KO @ 0h vs KO @ 18h
hist_4 <- ggplot(tc_data, aes(x = ko0ko18)) +
  geom_histogram(binwidth = 1, fill = "darkgoldenrod", color = "black") +
  labs(title = "Histogram of KO @ 0h vs KO @ 18h", x = "KO @ 0h vs KO @ 18h", y = "Count")

#Histogram of KO @ 0h vs WT @ 0h
hist_5 <- ggplot(tc_data, aes(x = ko0wt0)) +
  geom_histogram(binwidth = 1, fill = "darkorchid3", color = "black") +
  labs(title = "Histogram of KO @ 0h vs WT @ 0h", x = "KO @ 0h vs WT @ 0h", y = "Count")

#Histogram of KO @ 0h vs WT @ 6h
hist_6 <- ggplot(tc_data, aes(x = ko0wt6)) +
  geom_histogram(binwidth = 1, fill = "deepskyblue3", color = "black") +
  labs(title = "Histogram of KO @ 0h vs WT @ 6h", x = "KO @ 0h vs WT @ 6h", y = "Count")

#Histogram of KO @ 0h vs WT @ 18h
hist_7 <- ggplot(tc_data, aes(x = ko0wt18)) +
  geom_histogram(binwidth = 1, fill = "darkolivegreen", color = "black") +
  labs(title = "Histogram of KO @ 0h vs WT @ 18h", x = "KO @ 0h vs WT @ 18h", y = "Count")

grid.arrange(hist_1, hist_2, hist_3, hist_4, hist_5, hist_6, hist_7, ncol = 3)

```

--------------------
Volcano Plots
- Assess the relationship between magnitude of change and statistical significance of genes
- Large changes and higher significance towards the top and edges of the plot
--------------------
``` {r volcano}
#Volcano Plot of WT @ 0h vs WT @ 6h
vol1 <- ggplot(tc_data, aes(x = wt0wt6, y = -log10(padj_wt0wt6))) +
  geom_point(color = "aquamarine") +
  labs(title = "Volcano Plot of WT @ 0h vs WT @ 6h", x = "WT @ 0h vs WT @ 6h", y = "-log()")

#Volcano Plot of WT @ 0h vs WT @ 18h
vol2 <- ggplot(tc_data, aes(x = wt0wt18, y = -log10(padj_wt0wt18))) +
  geom_point(color = "coral") +
  labs(title = "Volcano Plot of WT @ 0h vs WT @ 18h", x = "WT @ 0h vs WT @ 18h", y = "-log()")

#Volcano Plot of KO @ 0h vs KO @ 6h
vol3 <- ggplot(tc_data, aes(x = ko0ko6, y = -log10(padj_ko0ko6))) +
  geom_point(color = "brown2") +
  labs(title = "Volcano Plot of KO @ 0h vs KO @ 6h", x = "KO @ 0h vs KO @ 6h", y = "-log()")

#Volcano Plot of KO @ 0h vs KO @ 18h
vol4 <- ggplot(tc_data, aes(x = ko0ko18, y = -log10(padj_ko0ko18))) +
  geom_point(color = "darkgoldenrod") +
  labs(title = "Volcano Plot of KO @ 0h vs KO @ 18h", x = "KO @ 0h vs KO @ 18h", y = "-log()")

#Volcano Plot of KO @ 0h vs WT @ 0h
vol5 <- ggplot(tc_data, aes(x = ko0wt0, y = -log10(padj_ko0wt0))) +
  geom_point(color = "darkorchid3") +
  labs(title = "Volcano Plot of KO @ 0h vs WT @ 0h", x = "KO @ 0h vs WT @ 0h", y = "-log()")

#Volcano Plot of KO @ 0h vs WT @ 6h
vol6 <- ggplot(tc_data, aes(x = ko0wt6, y = -log10(padj_ko0wt6))) +
  geom_point(color = "deepskyblue3") +
  labs(title = "Volcano Plot of KO @ 0h vs WT @ 6h", x = "KO @ 0h vs WT @ 6h", y = "-log()")

#Volcano Plot of KO @ 0h vs WT @ 18h
vol7 <- ggplot(tc_data, aes(x = ko0wt18, y = -log10(padj_ko0wt18))) +
  geom_point(color = "darkolivegreen") +
  labs(title = "Volcano Plot of KO @ 0h vs WT @ 18h", x = "KO @ 0h vs WT @ 18h", y = "-log()")

grid.arrange(vol1, vol2, vol3, vol4, vol5, vol6, vol7, ncol = 3)

```

--------------------
Significant Magnitudes of Change
--------------------
``` {r significant}
#Threshold
padj_threshold <- 0.05

#Significant for all comparisons
sig_all <- tc_data %>%
  filter(
    padj_wt0wt6 < padj_threshold &
    padj_wt0wt18 < padj_threshold &
    padj_ko0ko6 < padj_threshold &
    padj_ko0ko18 < padj_threshold &
    padj_ko0wt0 < padj_threshold &
    padj_ko0wt6 < padj_threshold &
    padj_ko0wt18 < padj_threshold
  )

#Significant for all wt vs wt comparisons
sig_wt <- tc_data %>%
  filter(
    padj_wt0wt6 < padj_threshold &
    padj_wt0wt18 < padj_threshold 
  )

#Significant for all ko vs ko comparisons
sig_ko <- tc_data %>%
  filter(
    padj_ko0ko6 < padj_threshold &
    padj_ko0ko18 < padj_threshold 
  )

#Significant for all ko vs wt comparisons
sig_wtko <- tc_data %>%
  filter(
    padj_ko0wt0 < padj_threshold &
    padj_ko0wt6 < padj_threshold &
    padj_ko0wt18 < padj_threshold
  )

#Individual significance
sig_wt0wt6 <- tc_data %>%
  filter(padj_wt0wt6 < padj_threshold)

sig_wt0wt18 <- tc_data %>%
  filter(padj_wt0wt18 < padj_threshold)

sig_ko0ko6 <- tc_data %>%
  filter(padj_ko0ko6 < padj_threshold)

sig_ko0ko18 <- tc_data %>%
  filter(padj_ko0ko18 < padj_threshold)

sig_ko0wt0 <- tc_data %>%
  filter(padj_ko0wt0 < padj_threshold)

sig_ko0wt6 <- tc_data %>%
  filter(padj_ko0wt6 < padj_threshold)

sig_ko0wt18 <- tc_data %>%
  filter(padj_ko0wt18 < padj_threshold)

```

``` {r compare}
# Keep only gene_names that have more than one gene_type
multi_type_genes <- tc_data %>%
  group_by(gene_name) %>%
  filter(n_distinct(gene_type) > 1) %>%
  ungroup() %>%
  arrange(gene_name)

# Pivot longer to compare across all timepoints
tc_long <- multi_type_genes %>%
  pivot_longer(
    cols = c(wt0wt6, wt0wt18, ko0ko6, ko0ko18, ko0wt0, ko0wt6, ko0wt18),
    names_to = "comparison",
    values_to = "logFC"
  )

# Function to apply ANOVA per gene_name & comparison
anova_results <- tc_long %>%
  group_by(gene_name, comparison) %>%
  filter(!is.na(logFC)) %>%
  filter(n_distinct(gene_type) > 1) %>%  # Ensure ≥2 gene types
  do({
    model <- aov(logFC ~ gene_type, data = .)
    tibble(
      p_value = summary(model)[[1]][["Pr(>F)"]][1]
    )
  }) %>%
  ungroup() %>%
  mutate(padj = p.adjust(p_value, method = "fdr"))


# Filter significant comparisons (e.g., FDR < 0.05)
sig_anova <- anova_results %>%
  filter(padj < 0.05)

# Example plot function
library(ggpubr)

plot_list <- list()

for (i in 1:nrow(sig_anova)) {
  gene <- sig_anova$gene_name[i]
  comp <- sig_anova$comparison[i]

  p <- tc_long %>%
    filter(gene_name == gene, comparison == comp) %>%
    ggplot(aes(x = gene_type, y = logFC, fill = gene_type)) +
    geom_boxplot() +
    geom_jitter(width = 0.1, alpha = 0.5) +
    labs(title = paste("Gene:", gene, "| Comparison:", comp),
         x = "Gene Type", y = "Log Fold Change") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  plot_list[[length(plot_list) + 1]] <- p
}

# Arrange plots in a grid (show first 6 as an example)
grid.arrange(grobs = plot_list[1:min(6, length(plot_list))], ncol = 2)


```

--------------------
Different things we can look at / Questions
- Can you predict what type of RNA or pseudo gene is being tested (column B) based off of the presence of a specific gene at comparison ___? [decision trees]
- Something to do with 0 values? No change... why? Similarities?
- Different time comparisons 0 vs 6 vs 18 by type of RNA/pseudo gene (which class/type has bigger differences)
--------------------




