---
title: "Vaccine"
author: "Summer"
date: "2025-07-01"
output: html_document
---

--------------------
Key
--------------------
NHS: Normal healthy subjects
ASYM: Chagas disease patients with no clinical symptoms
SYM: Chagas disease patients with clinical symptoms
ID: Identification marker for 1 of the 42 patients studied

--------------------
Definitions
--------------------
Serum: Liquid secreted after blood clots
Plasma: Liquid secreted when anti-coagulant is applied to blood
Copeptin:
- Peptide derived from the precursor of vasopressin
- Serves as a surrogate biomarker for vasopressin release
- Useful id diagnosing diabetes insipidus, heart failure, and acute myocardial infarction since vasopression is unstable in blood but copeptin is not
HnRNPA1:
- RNA binding protein involved in pre-mRNA splicing, transport, and stability
- Regulates gene expression post-transcriptionally
- Mutations associated with neurodegenerative diseases
Endostatin:
- Natural inhibtor of angiogensis
- Blocks formation of new blood vessels (tumor suppression)
Myostatin:
- Inhibits muscle growth
- Mutations that inactivate myostatin lead to muscle hypertrophy
8-OhDG:
- Marker of oxidative DNA damage
- Formed when DNA is attacked by reactive oxygen species
- Elevated levels seen in cancer, neurodegenerative diseases, diabetes, and agings. Assess oxidative stress
PARP1:
- DNA repair enzyme
- Detects DNA strand breaks and helps repair them by adding ADP-ribose chains to itself and other proteins

--------------------
Install important packages and libraries
--------------------
```{r package}
library(dplyr)
library(readxl)
library(corrplot)
library(ggplot2)
library(tidyverse)
library(emmeans)
library(ggpubr)
library(gridExtra)
library(tidyr)
library(broom)
library(ggrepel)
library(writexl)
library(rlang)
```

--------------------
Load data from excel
--------------------
```{r load}
file_path <- ("/Users/summerjohnson/Documents/GitHub/UTMB-SIBDS/proteinparam.xlsx")
data <- read_xlsx("/Users/summerjohnson/Documents/GitHub/UTMB-SIBDS/proteinparam.xlsx")
head(data)
```